---
title: "Modelling the decline in Listeria Cases"
subtitle: "Initial concept note - Confidential"
author: "Prepared by SACEMA"
output: 
  pdf_document:
    fig_caption: yes
    highlight: tango
    latex_engine: xelatex
  html_document: default
mainfont: Arial
geometry: margin=.75in
fontsize: 12pt
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
require(tidyverse)
require(bbmle)
source('epiCurve.R')
#source('incubationDist.R')
sampleDelay <- 3
bsIterations <- 200
source('incubationDistSens.R') # Assumes samples are acquired within sampleDelay days of onset
#source('incubationDist.R')
source('epiFunctions.R')
baseline <- mean(cases$count)
recallEffect <- 0.65 # after 1 week
expectedShelfLife <- 2 # weeks
```

## Baseline cases

The expected baseline number of cases was set to the mean weekly number of cases for Week 40 of 2017 through week 8 of 2018 (n = `r round(baseline,1)`). Observed variation around this mean was approximately Poisson distributed, so Poisson noise around the expected value is assumed in the model.

## Decline in risky exposures

An initial assessment suggests that the product recall was `r recallEffect*100`\% effective within one week. We assume a corresponding decline in risky exposures over this timeframe. We assume a further geometric decline through time as the remaining contaminated product is consumed or discarded. We have initially assumed the average remaining time that the contaminated product poses a risk is `r round(expectedShelfLife,1)` weeks. This assumption is probably conservative and is easily modified. The resulting temporal profile for risky exposures is represetnted below (Figure 1A).

```{r exposure, echo=FALSE}
ep <- exposureProfile(baseline,recallEffect)
minGen <- qpois(0.025,ep$riskExp)
maxGen <- qpois(0.975,ep$riskExp)
```

## Anticipated future cases

The number of future cases is determined by two primary factors: the temporal profile of risky exposures and the incubation period. The incubation period distribution has been characterized, initially, based on data from previous outbreaks in Europe and the US (see Goulet _et al_. 2013 _BMC Infectious Diseases_, Figure 1B). As data on the incubation period becomes available for the current outbreak in South Africa, we will use these data to update estimates that are likely to be more relevant for this outbreak, given the differences in background population and the source of exposure. Note that the long tail in the distribution of pregnacy-related cases is not particularly well captured by the iniital characterization of the incubation period distribution.

```{r,echo=FALSE,fig.width=7.5, fig.height=3, fig.cap='(A) Temporal profile of risky expsoures. (B) Comparison of non-parametric (grey) and parametric (green) survival analyses for incubation data. Based on data from Goulet 2013.'}
par(mar=c(5,4,2,2), mfcol=c(1,2))
plot(ep$week+.5,ep$riskExp,ylim=c(0,45),bty='L',ann=F,yaxt='n',xaxt='n',yaxs='i',xpd=T,type='s',lwd=3)
lines(ep$week+.5,minGen,col='darkgrey',type='s')
lines(ep$week+.5,maxGen,col='darkgrey',type='s')
axis(2,seq(0,45,5),cex.axis=.85)
axis(1,seq(1,18,2),cex.axis=.85)
mtext('Week of the year',1,3)
mtext('Cases generated',2,2.5)
mtext('A',3,line=1,adj=-.3,font=2)

par(mar=c(5,4,2,1))
plot(npmle_perInc,ylim=c(0,1),bty='L',ann=F,yaxt='n',yaxs='i',xpd=NULL)
axis(2,seq(0,1,.4),cex.axis=.85)
axis(2,seq(.2,1,.4),cex.axis=.85)
mtext('Time since infection (days)',1,3)
mtext('Proportion incubating',2,2.5)
mtext('B',3,line=1,adj=-.3,font=2)
curve(1-pgamma(x,meanIncGamma*rateIncGamma,rateIncGamma),add=T,lwd=3,col='darkgreen')
points(meanIncGamma,1.1,pch=21,cex=1.1,xpd=T,col='darkgreen',lwd=3)
arrows(ciIncGamma[1],1.1,ciIncGamma[2],1.1,le=.02,angle = 90,lwd=3,code = 3,col='darkgreen',xpd=T)
mtext('Mean and 95% CI',3,line=.5,adj=1.1,font=2,col='darkgreen',cex=.8,xpd=NULL)
abline(v=3.5+seq(0,70,7),col='red',lty=3)
```

Combining the temporal profile for risky exposures and the estimated distribution of the incubation period, we can begin to assess the anticipated decline in the epidemic curve as a result of increased awareness and product recall. Figure 2 shows plausible ranges for the number of cases under the assumptions of the model.

```{r,echo=FALSE,fig.width=4, fig.height=4, fig.cap='Range of cases expected through time, from Week 8 to Week 18 of 2018.'}
weekRange <- 8:18
par(mar=c(5,4,2,2))
plot(weekRange,rep(NA,11),ylim=c(0,45),bty='L',xlab='Week of the year',ylab='Anticipated number of cases')
arrows(weekRange,projCurve(minGen)[weekRange],weekRange,projCurve(maxGen)[weekRange],le=.02,angle = 90,lwd=3,code = 3,col='darkblue',xpd=T)
#projCurve(ep$riskExp)
# Can we get to probablility of seeing no cases for two consecutive weeks through time?
```

**If the number of cases through time is consistently or dramatically above the maximum of the expected range, this indicates that the assumptions of the model are violated in important ways and is probably a sign that there is an additonal source of risky exposure that remains unidentified.**





**Contact:** Please address questions, feedback, and requests related to this document to Prof. Juliet Pulliam, Director of SACEMA, at <pulliam@sun.ac.za>.